Extending statistical models for source attribution of zoonotic diseases: A study of campylobacteriosis

### 2018-02-05 paper/finalish/article_1023.tex
add simple changes to first 3 sections

\begin{document}
\runninghead{Extending statistical models for the source attribution of zoonotic diseases}
\title{Extending statistical models for the source attribution of zoonotic diseases: A study of campylobacteriosis} 
\author{Sih-Jing Liao\affil{a},
Jonathan Marshall\affil{a}\corrauth\ \footnotemark[2], Martin L. Hazelton\affil{a} and Nigel P. French\affil{b}}
\address{\affilnum{a} Institute of Fundamental Sciences-Statistics, Massey University, Palmerston North, New Zealand\\
\affilnum{b} mEpiLab, Hopkirk Research Institute, Massey University, Palmerston North, New Zealand}
\corraddr{Jonathan Marshall, Institute of Fundamental Sciences-Statistics, Massey University, Private Bag 11222, Palmerston North, New Zealand\\ \footnotemark[2] E-mail: J.C.marshall@massey.ac.nz}

@@

Preventing and controlling zoonoses through the design and implementation of public health policies requires a thorough understanding of transmission pathways. Modeling jointly the epidemiological data and genetic information on the microbial isolates derived from cases provides a methodology for tracing back the source of infection. The isolate genetic information and the epidemiological data may be modeled at a variety of levels. In this paper, the attribution probability for human cases of campylobacteriosis for each source, conditional on the extent to which each case resides in a rural compared to urban environment, is estimated. A model that incorporates genetic data and evolutionary processes is applied alongside a newly developed genetic data-free model. We show that the inference from each model is comparable, and that the effect of ``rurality'' may be modeled linearly, with increasing rurality leading to increasing likelihood of ruminant-sourced campylobacteriosis.

@@

\section{Introduction}
Modeling of disease surveillance data to explore patterns of infectious diseases has had a long history in public health. Infectious diseases can cause high economic and medical costs due to morbidity and mortality. In recent decades, the annual number of global deaths caused by infections has leveled off at approximate 15 million and may remain at this level for the next three decades \cite{DyeC, WHOM}. In order for such an enormous health burden to be reduced, preventing and controlling infectious diseases becomes extraordinarily important, which depends on how much we know about the nature of disease transmission. For zoonotic diseases, where transmission to humans from animal reservoirs may be via food, water, through environmental contamination or via direct contact with animals, knowledge of the potential sources and pathways of infection is key to reducing the burden of disease. Quantifying the burden of illness attributable to each pathway and source, provides valuable information to health professionals making decisions on public health policy for disease control.

@@

The first step for attribution models that utilize genetic information is building the sampling distribution of genotypes among each putative source. This may range from using the proportion of each observed genotype \cite{Hald} on each source through to utilizing allelic profile information to derive mutation and recombination rates within each source, and migration rates between each source \cite{Wilso}. A key question is whether more complex genetic models yield superior attribution results or whether a significantly simpler model may suffice, however few authors in the literature have addressed this point. This becomes more important as model complexity extends to include epidemiological covariates. We are therefore motivated to develop a simple model in order to assess the additional information that the more complex models provide.

In this paper we develop statistical models for the source attribution, and use a case study of human campylobacteriosis conducted in New Zealand \cite{Marsh} as a demonstration. Human campylobacteriosis is caused mainly by \textit{C. jejuni} and \textit{C. coli} which are the dominant species associated with 80\% and approximately 15\% of illnesses, respectively \cite{Guert}. The common symptoms of infection are diarrhea, abdominal pain and fever. However, a severe complication named Guillain Barre syndrome may develop, which is a life-threatening disease that weakens the nervous system and leads to paralysis in limbs and respiration \cite{Hanha}. The pathogen can be spread between animals, or from animals and wild birds to humans. The transmission route can be via drinking contaminated water, eating undercooked animal food products, or handling animal food products that are already contaminated by faeces. In this study, we compare the performance of the asymmetric Island model \cite{Wilso}, which considers genetic evolution when estimating the genotype sampling distribution on each source, to a simple model that utilizes only the prevalence of each type to derive the sampling distribution. We then extend both models to incorporate covariates such as rurality indicating the area where a human case resides, exploring the effect of human case rurality on attribution results, modeling attribution by rurality using a linear trend or using separate categories in a Bayesian context, and performing model comparison.

\section{Models for source attribution with and without microbial genetic information}
\subsection{Notation and assumptions}
The ultimate goal of attribution models is to estimate the probability that the observed human cases arise from each putative source. Given genotyping information, we first estimate the sampling distribution of genotypes for each source, and then estimate the appropriate combinations of those genotype distributions that most likely give rise to that observed among human cases. Specifying first the sampling distribution of genotypes found on sources is fundamental for the purpose of not only exploring how it affects the source attribution probability, but also investigating the difference in attribution effect made between the model with genetic information and the model without it.

Suppose we have isolates collected from human and non-human cases, of which $H$ isolates belong to humans, and the remaining $N$ isolates are categorized in $J$ groups as the major sources attributed to the infection. Let $I$ genotypes be the total number of unique STs detected from all isolates, and denote $n_j$ as the marginal frequency of STs found in source $j$, where $\sum_j^J n_j = N$. Typically, the number of detected STs $I$ is smaller than the sample size of isolates as multiple isolates will be of the same type.

The likelihood of observing human cases with types $\text{ST}_1, \text{ST}_2, \ldots \text{ST}_H$ follows a multinomial distribution, which via the law of total probability may be expressed as

\begin{align}
  L\Big(\text{ST}_1, \text{ST}_2, \cdots, \text{ST}_H \Big)=\prod_{h=1}^{H}\sum_{j=1}^{J} p\Big(\text{ST}_h \vert \text{source } j\Big) p\Big(\text{source }j\Big),
  \label{Olikh}
\end{align}

where $p(\text{ST}_h \vert \text{source }j)$ is the probability that $\text{ST}_h$ arises from the sampling distribution of source $j$, $p(\text{source }j)$ is the attribution probability that a random human case is infected from source $j$. Given we know $p(\text{ST}_h \vert \text{source }j)$, estimation of $p(\text{source }j)$ may be found by optimising the likelihood \ref{Olikh}, for example using a Metropolis Hastings algorithm within a Bayesian context, with suitable priors on $p(\text{source }j)$.

@@

To discover the effect of incorporating genetic information at the allelic profile level as used in the asymmetric Island model, a simple model is developed for the genotype sampling distribution. With the assumption that the observed distribution of genotypes is representative of the true distribution, we model observed genotypes using a multinomial distribution. Let $x_{ij}$ denote the count of ST-$i$ found in source $j$ with the probability $\pi_{ij}$, where $i=1, \ldots, I$ and $j=1, \ldots, J$. To make inference about $\pi_{ij}$, the likelihood is of a multinomial form,

\begin{align}
        L(\boldsymbol{\pi}_{j} ; \boldsymbol{x}_{j}) & = \frac{n_j!}{\prod_{i=1}^{I} x_{ij}!} \prod_{i=1}^I \pi_{ij}^{x_{ij}}, \notag
\end{align}

where $x_{ij}$ can be $0$, indicating STs are yet observed on the sources; $n_j=\sum_{i=1}^I x_{ij}$ is the total count of ST found on source $j$ and $\pi_{ij}$ is subject to $\sum_{i=1}^I \pi_{ij} =1$, and $0 \leq \pi_{ij} \leq 1$. As the family of Dirichlet distributions is a conjugate pair for the multinomial distribution, assume the prior for $\boldsymbol{\pi}_{j}$ follows a Dirichlet density with parameters $\boldsymbol{\alpha}_{j}$,

\begin{align}
  p(\boldsymbol{\pi}_{j}) & \propto \prod_{i=1}^{I}\pi_{ij}^{\alpha_{ij}-1}. \notag
\end{align}

Then the posterior for $\boldsymbol{\pi}_{j}$ takes the form of a Dirichlet probability function with parameters $(\boldsymbol{\alpha}_{j}+\boldsymbol{x}_{j}-\boldsymbol{1})$,

\begin{align}
        p(\boldsymbol{\pi}_{j} \vert \boldsymbol{x}_{j} ) & \propto L(\boldsymbol{\pi}_{j} ; \boldsymbol{x}_{j}) p(\boldsymbol{\pi}_{j}) \notag \\ 
       & \propto \prod_{i=1}^{I}\pi_{ij}^{\alpha_{ij} + x_{ij}-1}, \quad \alpha_{ij} >0. \notag
 \end{align}

To express the belief that every isolate is equally likely \emph{apriori}, the parameter of the Dirichlet prior is assumed as $\boldsymbol{\alpha}_{j}=\boldsymbol{1}$. Therefore, $p(\text{ST}_h \vert \text{source }j)$, $\boldsymbol{X}$) can be obtained by simulating from the Dirichlet posterior.

Lastly, to estimate the attribution probability, 500 samples are taken from $p(\text{ST}_h \vert \text{source }j)$ generated by the asymmetric Island or simple models. For each posterior sample, we estimate $p(\text{source }j)$ and its uncertainty from the attribution model~(\ref{Olikh}). This has the effect of integrating over the uncertainty in $p(\text{ST}_h \vert \text{source }j)$ when estimating $p(\text{source }j)$.

@@

The source attribution study described here uses MLST data collected in the Manawatu region of New Zealand from March, 2005 to December, 2014. The sampling was carried out over the same time period and from the same geographical location, providing 3,480 isolates taken from human and non-human cases, 1,400 of which are from humans. The non-human isolates are sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, we therefore put them into four groups representing major sources of infection: poultry, ruminants, water and others. The total number of unique STs typed from all isolates is 355, and 35\% of genotypes are found among human cases. Table~(\ref{tab1}) lists five common genotypes found in human and source isolates. The first four genotypes are frequently observed in human cases. For four sources, ST-45 and ST-474 are commonly detected in poultry, while ST-42 and ST-61 are typically found in ruminants. Some studies also discover that human cases with ST-45 or ST-474 are likely attributable to poultry, whereas human cases with ST-42 or ST-61 are more likely attributable to ruminants \cite{Muell, Coll, Cart}. Furthermore, it also shows that there are no observations of ST-2381 in humans, poultry or ruminants, indicating that the genotype may only appear in the water or wild birds.

The data also contain two location variables: one is binary, the other is categorical. The binary variable specifies whether a typed human case is from urban or rural areas, whereas the categorical variable further splits up the areas into seven levels. The levels are highly rural/remote area, rural area with low urban influence, rural area with moderate urban influence, rural area with high urban influence, independent urban area, satellite urban area and main urban area. Table~(\ref{tab2}) lists the number of typed human cases in each classification of rurality. The degree of rurality is a set of integers with a range of [-3,3] corresponding to each defined geographical level. Overall, more than 70\% of typed cases lived in urban areas, and approximately 7.5\% of individuals have no information about the location. We will neglect the missing data in the following analysis by assuming that these cases are randomly missing. Further, the table tells us that the population in 2006 and 2013 for each defined area \cite{Stat1, Stat2} is pretty similar, indicating that the population size in the region over the sampling period may be constant. We then calculate the infection rate per 100,000 population and illustrate the prevalence in urban and rural areas from 2006 to 2014 in Figure~(\ref{fig1}). The figure excludes the year of 2005 because the estimates population of both areas is unavailable and the data of typed human cases are incomplete for the entire year. As illustrated in the figure, the infection rates in 2006 are about 60 and 85 in urban and rural areas, respectively. We notice that the burden of infection in both areas drop remarkably in 2008, resulting from an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008. It shows the intervention improves the infection in urban areas, however, it only has a temporary effect in rural areas.

>>>>

\begin{document}
\runninghead{Extending statistical models for the source attribution of zoonotic diseases}
\title{Extending statistical models for source attribution of zoonotic diseases: A study of campylobacteriosis} 
\author{Sih-Jing Liao\affil{a},
Jonathan Marshall\affil{a}\corrauth\ \footnotemark[2], Martin L. Hazelton\affil{a} and Nigel P. French\affil{b}}
\address{\affilnum{a} Institute of Fundamental Sciences-Statistics, Massey University, Palmerston North, New Zealand\\
\affilnum{b} mEpiLab, Hopkirk Research Institute, Massey University, Palmerston North, New Zealand}
\corraddr{Jonathan Marshall, Institute of Fundamental Sciences-Statistics, Massey University, Private Bag 11222, Palmerston North, New Zealand\\ \footnotemark[2] E-mail: J.C.marshall@massey.ac.nz}

@@

Preventing and controlling zoonoses through the design and implementation of public health policies requires a thorough understanding of transmission pathways. Modeling jointly the epidemiological data and genetic information on the microbial isolates derived from cases provides a methodology for tracing back the source of infection. The isolate genetic information and the epidemiological data may be modeled at a variety of levels. In this paper, the attribution probability for human cases of campylobacteriosis for each source, conditional on the extent to which each case resides in a rural compared to urban environment, is estimated. A model that incorporates genetic data and evolutionary processes is applied alongside a newly developed genetic data-free model. We show that the inference from each model is comparable except for rare microbial genotypes, and that the effect of ``rurality'' may be modeled linearly on the logit scale, with increasing rurality leading to increasing likelihood of ruminant-sourced campylobacteriosis.

@@

\section{Introduction}
Modeling of disease surveillance data to explore patterns of infectious diseases has had a long history in public health. Infectious diseases can cause high economic and medical costs due to morbidity and mortality. In recent decades, the annual number of global deaths caused by infections has leveled off at approximate 15 million and may remain at this level for the next three decades \cite{DyeC, WHOM}. In order for such an enormous health burden to be reduced, preventing and controlling infectious diseases becomes extraordinarily important, which depends on how much we know about the nature of disease transmission. For zoonotic diseases, where transmission to humans from animal reservoirs may be via food, water, through environmental contamination or via direct contact with animals, knowledge of the potential sources and pathways of infection is key to reducing the burden of disease. Quantifying the burden of illness attributable to each pathway and source provides valuable information to health professionals making decisions on public health policy for disease control.

@@

The first step for attribution models that utilize genetic information is building the sampling distribution of genotypes among each putative source. This may range from using the proportion of each observed genotype \cite{Hald} on each source through to utilizing allelic profile information to derive mutation and recombination rates within each source, and migration rates between each source \cite{Wilso}. A key question is whether more complex genetic models yield superior attribution results or whether a significantly simpler model may suffice, but few authors in the literature have addressed this point. This becomes more important as model complexity extends to include epidemiological covariates. We are therefore motivated to develop a simple model in order to assess the additional information that the more complex models provide.

In this paper we develop statistical models for the source attribution, and use a case study of human campylobacteriosis conducted in New Zealand \cite{Marsh} as a demonstration. Human campylobacteriosis is caused mainly by \textit{C. jejuni} and \textit{C. coli} which are the dominant species associated with 80\% and approximately 15\% of illnesses, respectively \cite{Guert}. The common symptoms of infection are diarrhea, abdominal pain and fever. However, a severe complication named Guillain Barre syndrome may develop, which is a life-threatening disease that weakens the nervous system and leads to paralysis in limbs and respiration \cite{Hanha}. The pathogen can be spread between animals, or from animals and wild birds to humans. The transmission route can be via drinking contaminated water, eating undercooked animal food products, or handling animal food products that are already contaminated by faeces. In this study, we compare the performance of the asymmetric Island model \cite{Wilso}, which considers genetic evolution when estimating the genotype sampling distribution on each source, to a simple model that utilizes only the prevalence of each type to derive the sampling distribution. We then extend both models to incorporate covariates such as rurality indicating the area where a human case resides, exploring the effect of human case rurality on attribution results, modeling attribution by rurality using a linear trend on the logit scale or using separate categories in a Bayesian context, and performing model comparison.

\section{Models for source attribution with and without microbial genetic information}
\subsection{Notation and assumptions}
The ultimate goal of attribution models is to estimate the probability that the observed human cases arise from each putative source. Our data comprise microbial genotype information from each observed human case obtained from analysis of stool samples, and also from a pool of non-human cases corresponding to potential zoonotic sources of disease

Given this genotyping information, we first estimate the sampling distribution of genotypes for each source, and then estimate the appropriate combinations of those genotype distributions that most likely give rise to the set of genotypes observed among human cases. Specifying first the sampling distribution of genotypes found on sources is fundamental for the purpose of not only exploring how it affects the source attribution probability, but also investigating the difference in attribution effect made between the model with genetic information and the model without it.

Suppose we have isolates collected from human and non-human cases, of which $H$ isolates belong to humans, and the remaining $N$ isolates are categorized in $J$ groups as the major sources attributed to the infection. Let $I$ genotypes be the total number of unique sequence types (STs) detected from all isolates, and denote $n_j$ as the marginal frequency of STs found in source $j$, where $\sum_j^J n_j = N$. Typically, the number of detected STs $I$ is smaller than the sample size of isolates as multiple isolates will be of the same type.

The likelihood of observing human cases with types $\text{ST}_1, \text{ST}_2, \ldots \text{ST}_H$ follows a multinomial distribution, which via the law of total probability may be expressed as
\begin{align}
  L\Big(\text{ST}_1, \text{ST}_2, \cdots, \text{ST}_H \Big)=\prod_{h=1}^{H}\sum_{j=1}^{J} p\Big(\text{ST}_h \vert \text{source } j\Big) p\Big(\text{source }j\Big),
  \label{Olikh}
\end{align}
where $p(\text{ST}_h \vert \text{source }j)$ is the probability that $\text{ST}_h$ arises from the sampling distribution of source $j$, and $p(\text{source }j)$ is the attribution probability that a random human case is infected from source $j$. Given we know $p(\text{ST}_h \vert \text{source }j)$, estimation of $p(\text{source }j)$ may be found by optimising the likelihood \eqref{Olikh}, for example using a Metropolis Hastings algorithm within a Bayesian context, with suitable priors on $p(\text{source }j)$.

@@

To discover the effect of incorporating genetic information at the allelic profile level as used in the asymmetric Island model, a simple model is developed for the genotype sampling distribution. With the assumption that the observed distribution of genotypes is representative of the true distribution, we model observed genotypes using a multinomial distribution. Let $x_{ij}$ denote the count of ST-$i$ found in source $j$ with the probability $\pi_{ij}$, where $i=1, \ldots, I$ and $j=1, \ldots, J$. To make inference about $\pi_{ij}$, the likelihood is of a multinomial form,
\begin{align}
        L(\boldsymbol{\pi}_{j} ; \boldsymbol{x}_{j}) & = \frac{n_j!}{\prod_{i=1}^{I} x_{ij}!} \prod_{i=1}^I \pi_{ij}^{x_{ij}}, \notag
\end{align}
where $x_{ij}$ can be $0$, indicating STs are yet observed on the sources; $n_j=\sum_{i=1}^I x_{ij}$ is the total count of ST found on source $j$ and $\pi_{ij}$ is subject to $\sum_{i=1}^I \pi_{ij} =1$, and $0 \leq \pi_{ij} \leq 1$. As the family of Dirichlet distributions is a conjugate pair for the multinomial distribution, assume the prior for $\boldsymbol{\pi}_{j}$ follows a Dirichlet density with parameters $\boldsymbol{\alpha}_{j}$,
\begin{align}
  p(\boldsymbol{\pi}_{j}) & \propto \prod_{i=1}^{I}\pi_{ij}^{\alpha_{ij}-1}. \notag
\end{align}
Then the posterior for $\boldsymbol{\pi}_{j}$ takes the form of a Dirichlet probability function with parameters $(\boldsymbol{\alpha}_{j}+\boldsymbol{x}_{j}-\boldsymbol{1})$,
\begin{align}
        p(\boldsymbol{\pi}_{j} \vert \boldsymbol{x}_{j} ) & \propto L(\boldsymbol{\pi}_{j} ; \boldsymbol{x}_{j}) p(\boldsymbol{\pi}_{j}) \notag \\ 
       & \propto \prod_{i=1}^{I}\pi_{ij}^{\alpha_{ij} + x_{ij}-1}, \quad \alpha_{ij} >0. \notag
 \end{align}
To express the belief that every isolate is equally likely \emph{apriori}, the parameter of the Dirichlet prior is assumed as $\boldsymbol{\alpha}_{j}=\boldsymbol{1}$. Therefore, $p(\text{ST}_h \vert \text{source }j)$, $\boldsymbol{X}$) can be obtained by simulating from the Dirichlet posterior.

Lastly, to estimate the attribution probability, 500 samples are taken from $p(\text{ST}_h \vert \text{source }j)$ generated by the asymmetric Island or simple models. For each posterior sample, we make inferences for $p(\text{source }j)$ using the likelihood~\eqref{Olikh}. This has the effect of integrating over the uncertainty in $p(\text{ST}_h \vert \text{source }j)$ when estimating $p(\text{source }j)$.

@@

The source attribution study described here uses MLST data collected in the Manawatu region of New Zealand from March, 2005 to December, 2014. The sampling was carried out over the same time period and from the same geographical location, providing $H+N=3,480$ isolates taken from human and non-human cases, $H=1,400$ of which are from humans. The non-human isolates are sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, we therefore put them into four groups representing major sources of infection: poultry, ruminants, water and others. The total number of unique STs typed from all isolates is $I=355$, and 35\% of genotypes are found among human cases. Table~(\ref{tab1}) lists five common genotypes found in human and source isolates. The first four genotypes are frequently observed in human cases. For four sources, ST-45 and ST-474 are commonly detected in poultry, while ST-42 and ST-61 are typically found in ruminants. Some studies also discover that human cases with ST-45 or ST-474 are likely attributable to poultry, whereas human cases with ST-42 or ST-61 are more likely attributable to ruminants \cite{Muell, Coll, Cart}. Furthermore, it also shows that there are no observations of ST-2381 in humans, poultry or ruminants, indicating that the genotype may only appear in the water or wild birds.

The data also contain two location variables: one is binary, the other is categorical. The binary variable specifies whether a typed human case is from urban or rural areas, whereas the categorical variable further splits up the areas into seven levels. The levels are highly rural/remote area, rural area with low urban influence, rural area with moderate urban influence, rural area with high urban influence, independent urban area, satellite urban area and main urban area. Table~(\ref{tab2}) lists the number of typed human cases in each classification of rurality. The degree of rurality is a set of integers with a range of [-3,3] corresponding to each defined geographical level. Overall, more than 70\% of typed cases lived in urban areas, and approximately 7.5\% of individuals have no information about the location. We will neglect the missing data in the following analysis by assuming that these cases are randomly missing. Further, the table tells us that the population in 2006 and 2013 for each defined area \cite{Stat1, Stat2} is quite stable. We then calculate the infection rate per 100,000 population and illustrate the prevalence in urban and rural areas from 2006 to 2014 in Figure~(\ref{fig1}). The Figure excludes the year of 2005 because the estimates population of both areas is unavailable and the data of typed human cases are incomplete for the entire year. As illustrated in the Figure, the infection rates in 2006 are about 60 and 85 in urban and rural areas, respectively. We notice that the burden of infection in both areas drop remarkably in 2008, resulting from an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008. It shows the intervention improved infection rates in urban areas, however, it only has a temporary effect in rural areas.
